---
title: "tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(smartcount)
```

# Introduction
The `smartcount` package provides a streamlined workflow for count regression analysis. 

# Dataset
We use `bioChemists` dataset from the `pscl` package, which records the number of articles published by 915 PhD biochemists.

```{r}
library(pscl)
data("bioChemists")
head(bioChemists)
```

# Summarize Data
## Before fitting model, we examine whether the data user provides is a data frame and whether the response variable exists
```{r}
summarize_count_data(bioChemists, "art")

```

# Fit a Poisson Model
## We start with a Poisson regression
```{r}
fit_p <- fit_count(bioChemists, art~ fem + mar + kid5 + phd + ment, 
              model="poisson")
```

# Check Conditions
```{r}
check_conditions(fit_p)

```

# Try fit ZIP based on suggestions

# Evaluate and Comapre Models

# Interpret Final Model

# Conclusion



