---
title: "tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(smartcount)
```

# Introduction
The `smartcount` package provides a streamlined workflow for count regression analysis. 

# Dataset
We use `bioChemists` dataset from the `pscl` package, which records the number of articles published by 915 PhD biochemists.

```{r}
library(pscl)
data("bioChemists")
head(bioChemists)
```

# Summarize Data
## Before fitting model, we examine whether the data user provides is a data frame and whether the response variable exists
```{r}
summarize_count_data(bioChemists, "art")

```

# Fit a Poisson Model
## We start with a Poisson regression
```{r}
fit_p <- fit_count(bioChemists, art~ fem + mar + kid5 + phd + ment, 
              model="poisson")
```

# Check Conditions
```{r}
check_conditions(fit_p)
```

# Try fit ZIP based on suggestions

The Poisson check above suggested that the data may have overdispersion and/or zero inflation. We try a Zero-Inflated Poisson (ZIP) model to better 
handle the excess zeros.

```{r}
fit_zip <- fit_count(bioChemists, art ~ fem + mar + kid5 + phd + ment, 
                     model = "zip")
check_conditions(fit_zip)
```


# Evaluate and Compare Models

We use `evaluate_model()` to look at fit statistics. AIC and BIC measure
relative fit quality (lower is better).

```{r}
evaluate_model(fit_p)$stats
evaluate_model(fit_zip)$stats
```

To compare the two models directly, we pass both into `evaluate_model()`.
Since one model is zero-inflated and the other is not, the function
automatically uses the **Vuong test** for non-nested comparison.

```{r}
comparison <- evaluate_model(fit_p, fit_zip)
comparison$comparison
```

A negative Vuong z-statistic indicates that the second model (ZIP) is
preferred over the first (Poisson).


# Interpret Final Model

We use `interpret()` to translate the ZIP coefficients into clear,
non-technical language. The output table includes:

- `estimated_log`: coefficient on the log scale
- `estimated_exp`: exponentiated coefficient (rate ratio or odds ratio)
- `percentage`: percent change in expected count
- `ci_lower_pct` / `ci_upper_pct`: 95% confidence interval (percent scale)
- `z_value` / `p_value`: hypothesis test statistics
- `interpretation`: plain-English interpretation

```{r}
result <- interpret(fit_zip)
result[, c("variable", "percentage", "ci_lower_pct", "ci_upper_pct", "p_value")]
```

The full interpretation column:

```{r}
result$interpretation
```

ZIP coefficients are split into two parts:

- `count_xxx`: how the predictor affects the expected count among
  non-structural-zero observations
- `zero_xxx`: how the predictor affects the log-odds of being a
  structural zero (i.e., someone who never publishes)


# Extension: Generalized Poisson with Real Data

Beyond the standard count regression models, `smartcount` also supports
Generalized Poisson regression, which is more flexible than Poisson when
the data exhibit overdispersion (or, less commonly, underdispersion).

To demonstrate, we replicate the analysis from Akter et al. (2025),
*"Accounting for overdispersion in skilled antenatal care: Identifying
determinants using Bangladesh Demographic and Health Survey 2022 data"*
(PLOS ONE). The authors use a Generalized Poisson model to study factors
influencing the number of skilled antenatal care visits among 3,839
Bangladeshi women.

```{r}
df <- read.csv("../data-raw/bangladesh_anc.csv")

fit_gp <- fit_count(
  df,
  ANC_skilled ~ Mothers_Edu + Wealth_index + Media_Exposed,
  model = "genpois"
)

check_conditions(fit_gp)
```

The Pearson dispersion ratio close to 1 confirms that the Generalized
Poisson model handles the overdispersion well.

```{r}
interpret(fit_gp)
```

```{r}
evaluate_model(fit_gp)$stats
```


# Conclusion

The `smartcount` package provides an end-to-end workflow for count
regression analysis:

1. **Summarize** the data to understand its structure
2. **Fit** a starting Poisson model
3. **Check** the model's conditions
4. **Switch** to a more appropriate model if needed (Quasi-Poisson,
   Negative Binomial, ZIP, ZINB, or Generalized Poisson)
5. **Evaluate** model fit and compare alternatives
6. **Interpret** results in plain language

This integrated workflow saves users from manually stitching together
functions from different packages and applies appropriate diagnostic and
interpretation logic for each model type.


# References

- Akter et al. (2025). Accounting for overdispersion in skilled antenatal
  care: Identifying determinants using Bangladesh Demographic and Health
  Survey 2022 data. *PLOS ONE*.
  https://doi.org/10.1371/journal.pone.0346258

- Long, J. S. (1990). The origins of sex differences in science.
  *Social Forces*, 68(4), 1297-1316. (`bioChemists` dataset)



